---
output:
  github_document:
    html_preview: true
    toc: false
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # fig.path = "man/figures/README-",
  message = FALSE
)
```


# Installing the MS-DAP R package

This guide helps you install the MS-DAP R package and all of its software dependencies on your computer. If you are looking for the _Dockerized_ MS-DAP, [use this guide](docker.md).


## prequisites

**Windows**
```
You will need to install GIT, which is needed to automaticall retrieve R packages from github.com (may not be installed on your system). 
Download and install this tool, available at https://gitforwindows.org
```

**Ubuntu** (for Ubuntu 18.04 or older, see further https://github.com/ropensci/pdftools ):
```
sudo apt-get install git libnetcdf-dev netcdf-bin libpoppler-cpp-dev
```

**Fedora**
```
sudo dnf install git netcdf-devel netcdf poppler-cpp-devel
```

**MacOS**
first, install Homebrew if you haven't:  https://brew.sh
```
brew install git netcdf poppler automake
export PKG_CONFIG_PATH=/usr/local/Cellar/zlib/1.2.8/lib/pkgconfig:/usr/local/lib/pkgconfig:/opt/X11/lib/pkgconfig
```

**troubleshooting**

_poppler_ is a requirement for pdftools to function, which is used by MS-DAP to create PDF reports, if you have any trouble installing please check these resources:

- https://github.com/ropensci/pdftools
- https://github.com/politza/pdf-tools



## installing R (if you have not already)

- R version 4 is supported; 4.0.3 or newer is recommended.
- R version 3 is supported; 3.6.3 is recommended.

**Windows**

If you do not have R installed yet, go to https://cloud.r-project.org/bin/windows/

- download and install "base" R
- download and install "Rtools"

For reference, users that want to stick with R version 3 should grab R from here:

- base R 3.6.3 @ https://cloud.r-project.org/bin/windows/base/old/3.6.3/
- Rtools35 @ https://cran.r-project.org/bin/windows/Rtools/history.html

**Ubuntu**
```
sudo apt-get install r-base r-base-dev
```

**Fedora**
```
sudo yum install R
```

**MacOS**
- download and install _R-3.6.3.nn.pkg_ @ https://cran.r-project.org/bin/macosx/
- make sure to select Tcl/Tk and Textinfo during installation


## install RStudio Desktop

Download and install RStudio Desktop (free edition) from https://rstudio.com/products/rstudio/download/



## install the MS-DAP R package

Start RStudio (restart if it was already up-and-running to get a clean session) and run below code line-by-line so you can keep an eye on the results of each intermediate step in the RStudio console.

Notes on windows; 

1) When prompted to optionally compile packages from source you can simply select 'no' (to install a readily available binary instead). 
2) As noted during the TinyTex installation, the two popup errors stating _"The code execution cannot proceed because luatex.dll ..."_ can be safely dismissed.


```{r gh-installation, eval = FALSE}
### 1) setup required R packages
install.packages(c("devtools", "tidyverse", "tinytex", "BiocManager"))
# install LaTeX which is required to create the pdf report (you can safely dismiss both popup errors)
tinytex::install_tinytex()
# install packages from bioconductor
BiocManager::install(c('ProtGenerics', 'MSnbase', 'limma'), update=T, ask=F)

### 2) install MS-DAP R package directly from github
# first, make sure we don't halt on minor issues that can be ignored (eg; your R installation is a minor version behind)
Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS="true")
# install MS-DAP R package and all required dependencies (but don't check for updates on all packages, to minimize complexity)
devtools::install_github("ftwkoopmans/msdap", upgrade = "never")

# 3) if all went well, you can now load the msdap package.
library(msdap)

# If you get any warnings, no problem; you only need to be on the lookout for errors
```






**troubleshooting**

- if you encounter "installation path not writable" errors the user running R (/RStudio) may not have write access to the directory where the R packages are stored. Use the R command `.libPaths()` to print those locations. 
  - For windows users, the most robust workaround is to start a command prompt as administrator (start > type `cmd` > right-click `run as administrator`), open a commandline R session (eg; on the cmd prompt, type `"C:\Program Files\R\R-3.6.3\bin\R.exe"`) and retry your package installation.

- Further documentation for TinyTex is available at https://yihui.org/tinytex/
