---
output:
  github_document:
    html_preview: true
    toc: false
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # fig.path = "man/figures/README-",
  message = FALSE
)
```


# Installing the MS-DAP R package

This guide helps you install the MS-DAP R package and all of its software dependencies on your computer. If you are looking for the _Dockerized_ MS-DAP, [use this guide](docker.md).


## prequisites

You'll need to install the following system packages; netcdf and poppler

**Ubuntu** (for Ubuntu 18.04 or older, see further https://github.com/ropensci/pdftools ):
```
sudo apt-get install libnetcdf-dev netcdf-bin libpoppler-cpp-dev
```

**Fedora**
```
sudo dnf install netcdf-devel netcdf poppler-cpp-devel
```

**MacOS**
first, install Homebrew if you haven't:  https://brew.sh
```
brew install netcdf poppler automake
export PKG_CONFIG_PATH=/usr/local/Cellar/zlib/1.2.8/lib/pkgconfig:/usr/local/lib/pkgconfig:/opt/X11/lib/pkgconfig
```

**troubleshooting**

_poppler_ is a requirement for pdftools to function, which is used by MS-DAP to create PDF reports, if you have any trouble installing please check these resources:

- https://github.com/ropensci/pdftools
- https://github.com/politza/pdf-tools



## installing R (if you have not already)

An installation of R version 3.6 or newer is required, version 3.6.3 is recommended (4.0 is not supported yet, some of the packages we rely on are not compatible with 4.0 yet)

**Ubuntu**
```
sudo apt-get install r-base r-base-dev
```

**Fedora**
```
sudo yum install R
```

**MacOS**
- download and install _R-3.6.3.nn.pkg_ @ https://cran.r-project.org/bin/macosx/
- make sure to select Tcl/Tk and Textinfo during installation

**Windows**
- base R 3.6.3 @ https://cloud.r-project.org/bin/windows/base/old/3.6.3/
- Rtools35 @ https://cran.r-project.org/bin/windows/Rtools/history.html

**all systems**
finally, obtain RStudio Desktop (free edition) @ https://rstudio.com/products/rstudio/download/



## install the MS-DAP R package

Start RStudio (restart if it was already up-and-running to get a clean session) and run below code line-by-line so you can keep an eye on the results of each intermediate step in the RStudio console.

Notes on windows; 

1) when prompted to optionally compile packages from source you can simply select 'no' (to install a readily available binary instead). 
2) As noted during the TinyTex installation, the two popup errors stating _"The code execution cannot proceed because luatex.dll ..."_ can be safely dismissed.

```{r gh-installation, eval = FALSE}
# 1) install prequisites
install.packages(c("devtools", "tidyverse", "tinytex", "BiocManager"))
# install LaTeX which is required to create the QC pdf report (documentation/help available at https://yihui.org/tinytex/ )
tinytex::install_tinytex()
# install packages from bioconductor
BiocManager::install(c('ProtGenerics', 'MSnbase', 'limma'), update=T, ask=F)

# 2) install MS-DAP R package directly from github
# first, make sure we don't halt on minor issues that can be ignored (eg; your R installation is a minor version behind)
Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS="true")
# install MS-DAP R package and all required dependencies (but don't check for updates on all packages, to minimize complexity)
devtools::install_github("ftwkoopmans/msdap", upgrade = "never")

# 3) if all went well, you can now load the msdap package.
library(msdap)

# If you get any warnings, no problem, you only need to be on the lookout for errors
```


## Quickstart example

In this example we will analyze the Klaassen et al. (2018) dataset, assuming you have previously completed the installation steps above.

- To immediately view the expected results, [click here to download the PDF report](/examples/data/dataset_Klaassen2018_pmid26931375_report.pdf) we already prepared and uploaded.


1) <a id="raw-url" href="https://raw.githubusercontent.com/ftwkoopmans/msdap/master/examples/data/dataset_Klaassen2018_pmid26931375.zip">click here</a> to download the example dataset; a label-free WT/KO IP dataset @ Klaassen et al. 2018 (PMID 26931375) which we re-analyzed with MetaMorpheus.

2) unzip the files, for instance to `C:/data/`. In this example, your data would be located at `C:/data/dataset_Klaassen2018_pmid26931375` 

3) open RStudio and follow either the GUI or R-code quickstart guides below



### Quickstart: graphical user interface

Launching the GUI:

- start the MS-DAP graphical user interface (GUI) by running the following R command in the RStudio console at the bottom-left of the RStudio interface: `msdap::gui()` 

Data analysis steps:

- upstream software: _MetaMorpheus_
- input data: browse files > select _the folder where you extracted the dataset_ from the dropdown menu in the top-right > left-click the folder with the Klaassen et al. dataset in the _Directories_ panel > click select button on the bottom
- fasta: browse files > select both fasta files: left-click the first, hold shift, left-click the second fasta file (note that this shows the files from the most recent directory) > click select button on the bottom
- output directory: Optionally, change the location where the output files are stored. Note that MS-DAP results are stored in a subdirectory of the location you choose here.
- _next_
- load sample metadata table: double-click sample_metadata.xlsx
- _next_
- optionally, define a statistical contrast: drag and drop sample groups to 'A' and 'B' (or leave all groups at 'unused' to skip A/B testing)
- identified sample count: 1
- DEA algorithm: msqrob and msempire
- _start analysis_
- output log and result summary are shown in the GUI
- output files can be found in the _output directory_



### Quickstart: an example R script

Use a snippet of R code that uses the MS-DAP R package to analyse the Klaassen et al. (2018) dataset that was pre-processed with MetaMorpheus and included with MS-DAP:

- in the RStudio menu on the top-left; file > New File > R Script
- copy/paste the below code snippet to the panel in center of the screen
- importantly, adjust the file path in the `setwd()` command to the folder where you extracted the dataset!
- run all lines of code: select all (by mouse drag or click anywhere then `control+A`)  >  click the `run` button on top of the code panel to run all selected lines
- output log and result summary are shown at the bottom of the screen
- output files can be found in the folder defined in `setwd()`


```{r eval=FALSE}
library(msdap)

# !! importantly, adjust the below file path !!
# set working directory. use forward slashes in file paths
setwd("C:/data/dataset_Klaassen2018_pmid26931375")
# load the dataset
dataset = import_dataset_metamorpheus(path = getwd(), protein_qval_threshold = 0.05)
# gather protein information, such as gene symbols, from fasta files
dataset = import_fasta(dataset, files = c("UP000000589_10090.fasta", "UP000000589_10090_additional.fasta"))
# optionally, we can apply filters to protein names. here we remove keratin and IGGs from the IP dataset
dataset = remove_proteins_by_name(dataset, regular_expression = "ig \\S+ chain|keratin|GN=(krt|try|igk|igg|igkv|ighv|ighg)")
# import sample metadata and define a contrast for DEA
dataset = import_sample_metadata(dataset, filename = "sample_metadata.xlsx")
dataset = setup_contrasts(dataset, contrast_list = list(c("wt", "ko")))
# run the pipeline. check the documentation by running the R command: ?analysis_quickstart
dataset = analysis_quickstart(dataset,
                              filter_min_detect = 1,
                              filter_min_quant = 3,
                              filter_by_contrast = FALSE,
                              norm_algorithm = "msempire",
                              dea_algorithm = c("ebayes", "msempire", "msqrob"),
                              dea_qvalue_threshold = 0.05,
                              dea_log2foldchange_threshold = NA,
                              output_qc_report = TRUE,
                              # note; within the Docker container, /data maps to the directory on your computer
                              # defined when launching the MS-DAP Docker container
                              output_dir = getwd(),
                              output_within_timestamped_subdirectory = TRUE)
# optionally, print a brief summary to console
print_dataset_summary(dataset)
```




**troubleshooting**

- if you encounter "installation path not writable" errors the user running R (/RStudio) may not have write access to the directory where the R packages are stored. Use the R command `.libPaths()` to print those locations. 
  - For windows users, the most robust workaround is to start a command prompt as administrator (start > type `cmd` > right-click `run as administrator`), open a commandline R session (eg; on the cmd prompt, type `"C:\Program Files\R\R-3.6.3\bin\R.exe"`) and retry your package installation.
